OptiFit: an improved method for fitting amplicon sequences to existing OTUs

### 2021-10-24 paper/paper.Rmd
Minor edits / clarifications

To compare to the reference clustering methods used by QIIME2, we clustered each
dataset with VSEARCH against the Greengenes database of OTUs previously 
clustered at 97% sequence similarity.
Each reference OTU from the Greengenes 97% database contains one reference
sequence, and VSEARCH maps sequences to the reference based on each individual
query sequence's similarity to the single reference sequence.
In contrast, OptiFit accepts reference OTUs which each may contain multiple
sequences, and the sequence similarity between all query and reference sequences
is considered when assigning sequences to OTUs.
<!--**TODO: delete or move this sentence:** _De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
higher quality OTUs than VSEARCH in terms of MCC, but performed 
`r sec_opticlust_vs_vsearch`% slower than VSEARCH.-->
In closed reference mode, OptiFit produced `r mcc_closed_fit_db_vs_vsearch`%
higher quality OTUs than VSEARCH, but VSEARCH was able to cluster
`r frac_vsearch_vs_fit`% more query sequences than OptiFit to the Greengenes
reference database (Figure \ref{fig:results_sum}).
This is because VSEARCH only considers the distances between each query sequence
to the single reference sequence, while OptiFit considers the distances between
all pairs of reference and query sequences in an OTU. 
When open reference clustering, OptiFit produced higher quality OTUs than
VSEARCH against the Greengenes database, with median MCC scores of 
`r open_fit_gg_mcc` and `r open_vsearch_mcc`, respectively).
In terms of run time, OptiFit outperformed VSEARCH in both closed and open
reference mode by `r sec_closed_fit_db_vs_vsearch`% and 
`r sec_vsearch_vs_open_fit_db`% on average, respectively.
Thus, the more stringent OTU definition employed by OptiFit, which prefers the
query sequence to be similar to all other sequences in the OTU rather than to only one
sequence, resulted in fewer sequences being clustered to reference OTUs than when
using VSEARCH, but caused OptiFit to outperform VSEARCH in terms of both OTU
quality and execution time.

@@

To generate reference OTUs from independent databases, we downloaded sequences
from the Greengenes database (v13_8_99) [@desantis_greengenes_2006], SILVA
non-redundant database (v132) [@quast_silva_2013], and the Ribosomal Database
Project (v16) [@cole_ribosomal_2014].
These sequences were processed using the same steps outlined above followed by
clustering sequences into _de novo_ OTUs with OptiClust.
Processed reads from each of the four datasets were clustered with OptiFit to
the reference OTUs generated from each of the three databases.
When reference clustering with VSEARCH, processed datasets were clustered directly to
the unprocessed Greengenes 97% OTU reference alignment, since this method is how
VSEARCH is typically used by the QIIME2 software reference-based clustering
[@bolyen_reproducible_2019; @noauthor_clustering_nodate].

@@

For each dataset, a fraction of the sequences was selected to be clustered _de
novo_ into reference OTUs with OptiClust.
We used three methods for selecting the fraction of sequences to be used
as the reference: a simple random sample, weighting sequences by relative
abundance, and weighting by similarity to other sequences in the dataset.
Dataset splitting was repeated with reference fractions ranging from 10% to 90%
of the dataset and for 100 random seeds.
For each dataset split, the remaining query sequences were clustered into the
reference OTUs with OptiFit.

@@

We implemented the analysis workflow in Snakemake [@koster_snakemake_2012] and
wrote scripts in R [@r_core_team_r_2020], Python [@van_rossum_python_2009], and
GNU bash [@noauthor_bash_nodate].
Software used includes mothur v1.45.0 [@schloss_introducing_2009], VSEARCH
v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the tidyverse
metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], ggtext
[@wilke_ggtext_2020], the SRA toolkit [@noauthor_sra-tools_nodate], and the
conda environment manager [@noauthor_anaconda_2016].
The complete workflow, manuscript, and conda environment are available at
https://github.com/SchlossLab/Sovacool_OptiFit_2021.
<!-- TODO: update repo link with journal title -->

>>>>

To compare to the reference clustering methods used by QIIME2, we clustered each
dataset with VSEARCH against the Greengenes database of OTUs previously 
clustered at 97% sequence similarity.
Each reference OTU from the Greengenes 97% database contains one reference
sequence, and VSEARCH maps sequences to the reference based on each individual
query sequence's similarity to the single reference sequence.
In contrast, OptiFit accepts reference OTUs which each may contain multiple
sequences, and the sequence similarity between all query and reference sequences
is considered when assigning sequences to OTUs.
<!--**TODO: delete or move this sentence:** _De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
higher quality OTUs than VSEARCH in terms of MCC, but performed 
`r sec_opticlust_vs_vsearch`% slower than VSEARCH.-->
In closed reference mode, OptiFit produced `r mcc_closed_fit_db_vs_vsearch`%
higher quality OTUs than VSEARCH, but VSEARCH was able to cluster
`r frac_vsearch_vs_fit`% more query sequences than OptiFit to the Greengenes
reference database (Figure \ref{fig:results_sum}).
This is because VSEARCH only considers the distances between each query sequence
to the single reference sequence, while OptiFit considers the distances between
all pairs of reference and query sequences in an OTU. 
When open reference clustering, OptiFit produced higher quality OTUs than
VSEARCH against the Greengenes database, with median MCC scores of 
`r open_fit_gg_mcc` and `r open_vsearch_mcc`, respectively.
In terms of run time, OptiFit outperformed VSEARCH in both closed and open
reference mode by `r sec_closed_fit_db_vs_vsearch`% and 
`r sec_vsearch_vs_open_fit_db`% on average, respectively.
Thus, the more stringent OTU definition employed by OptiFit, which prefers the
query sequence to be similar to all other sequences in the OTU rather than to only one
sequence, resulted in fewer sequences being clustered to reference OTUs than when
using VSEARCH, but caused OptiFit to outperform VSEARCH in terms of both OTU
quality and execution time.

@@

To generate reference OTUs from public databases, we downloaded sequences
from the Greengenes database (v13_8_99) [@desantis_greengenes_2006], SILVA
non-redundant database (v132) [@quast_silva_2013], and the Ribosomal Database
Project (v16) [@cole_ribosomal_2014].
These sequences were processed using the same steps outlined above followed by
clustering sequences into _de novo_ OTUs with OptiClust.
Processed reads from each of the four datasets were clustered with OptiFit to
the reference OTUs generated from each of the three databases.
When reference clustering with VSEARCH, processed datasets were clustered directly to
the unprocessed Greengenes 97% OTU reference alignment, since this method is how
VSEARCH is typically used by the QIIME2 software for reference-based clustering
[@bolyen_reproducible_2019; @noauthor_clustering_nodate].

@@

For each dataset, half of the sequences were selected to be clustered _de
novo_ into reference OTUs with OptiClust.
We used three methods for selecting the subset of sequences to be used
as the reference: a simple random sample, weighting sequences by relative
abundance, and weighting by similarity to other sequences in the dataset.
Dataset splitting was repeated with 100 random seeds.
With the simple random sampling method, dataset splitting was also repeated with
reference fractions ranging from 10% to 90% of the dataset.
For each dataset split, the remaining query sequences were clustered into the
reference OTUs with OptiFit.

@@

We implemented the analysis workflow in Snakemake [@koster_snakemake_2012] 
and wrote scripts in R [@r_core_team_r_2020], 
Python [@van_rossum_python_2009], 
and GNU bash [@noauthor_bash_nodate].
Software used includes mothur v1.46.1 [@schloss_introducing_2009], 
VSEARCH v2.15.2 [@rognes_vsearch_2016], 
the tidyverse metapackage [@wickham_welcome_2019], 
R Markdown [@xie_r_2018], 
ggraph [@pederson_ggraph_2021], 
ggtext [@wilke_ggtext_2020], 
numpy [@harris_array_2020], 
the SRA toolkit [@noauthor_sra-tools_nodate], 
and conda [@noauthor_anaconda_2016].
The complete workflow and supporting files required to reproduce this manuscript
are available at https://github.com/SchlossLab/Sovacool_OptiFit_2021.
<!-- TODO: update repo link with journal title -->
